Estimating length of SARS-CoV-2 from doubly interval censored data with missed infections

### 2024-03-19 incidence-prevalence.tex
spelling / grammar check

I start by defining these terms in \cref{inc-prev:sec:definitions}.
I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process} and the \emph{observation process}.
The mathematical description of are an infection model, disease model, and observation model respectively.

@@

The prevalence process (\cref{inc-prev:sec:prevalence-process}) describes the relationship between incidence and prevalence.
Here, I adapt previous work to the SARS-CoV-2 context.
In particular, recovering from infection needs to be considering, which was not the case for HIV/AIDS.

@@

Time will be considered discrete in this chapter, except in \cref{SEIR:sec:mechanistic-models}.
I use days, however, the results generalise in the obvious way to other time units.
I choose discrete time for pragmatic reasons.
Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection); these variations are not of interest in most contexts.

\emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
The timing of infection events is very rarely directly observable and must be inferred from other data.
Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.

\emph{Prevalence} of a disease on day $t$, is the proportion of people with the disease on that day~\autocite[90]{lashModern}.
I refer to individuals with the disease as \emph{prevalent} individuals.
Whether an individual is prevalent is often expensive or error-prone to measure, therefore, it will often be regarded as a latent quantity.
Furthermore, what is meant by ``having'' a disease can depend on context; for example, it could be the presence of symptoms, the ability to transmit, or detectable status (the difference between these was discussed in \cref{E-biology-data:sec:natural-history})
Within this thesis, I define an individual as prevalent on day $t$ if they are detectable (detectable is defined in \cref{E-biology-data:sec:detectable}).
I assume that an individual's prevalent status cannot change over the course of a day.
Therefore, prevalence is constant over any single day $t$.
I denote the number of prevalent individuals on day $t$ as $P_t$.
Therefore, the prevalence is $P_t/\Npop$.

\emph{Duration}, or \emph{time-to-event}, distributions can be expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}).
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$.
$D_i = E_i - B_i + 1$ is their \emph{duration of positivity}, the number of days for which they are prevalent.
I model the $D_i$s as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and it will normally be expressed in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.

Since $D_i$ is a discrete random variable, $0 \leq f_D(d) \leq 1$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
This implies that $F_D(d) \in [0, 1]$ and increases monotonically.
Therefore, $S(d) \in [0, 1]$ and decreases monotonically.
Similarly, $\lambda(d) \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterisation for inference.

@@

A point process is a random function of continuous time $K(t)$.
$K(t)$ must take non-negative integer values and be monotonically increasing in $t$.
The interpretation of $K(t)$ is the number of events, here infections, that have occurred in the interval $(-\infty, t]$~\autocite[244]{yanDistribution}.
In the context of a disease, the point process can be discretised into the incidence on day $t$ as $Z_t = K(t) - K(t-1)$, where $t$ is an integer.
I will explain two approaches to modelling the infection process.

First, in \cref{inc-prev:sec:poisson-process}, as a \emph{Poisson process}.
A Poisson process is a widely-used type of counting process both within and outside of epidemiology.
It makes no direct assumptions on the underlying biological or epidemiological process, and is therefore a flexible model.

@@

A commonly-used point process for infectious disease epidemiology is the time-inhomogenous Poisson process~\autocite[e.g., in the context of HIV,][]{brookmeyerMethod,paganoHIV,rosenbergBackcalculation,brookmeyerBackcalculation}.
% A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
Extensions to this model, which I do not consider here, have incorporated covariates, such as geography or age~\autocite[e.g.][]{diggleModeling}.

@@

% A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
% A super-spreading event is an event where many more people than average are infected.
% That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
% Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocite{endoEstimating,adamClustering}{mccloskeySARS}.
% An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
% One infected individual is thought to have infected at least 32 further individuals during a single choir practice.
% For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
% Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t) \dist \Poi(\int_{t}^{s} \nu(r) dr)$ with $K(s) - K(t) \dist \NBc(\int_{t}^{s} \nu(r) dr, k)$ where $k$ is the overdispersion parameter, normally assumed to be constant absent interventions.
% Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterisations and moments).
% %The negative binomial tends to Poisson as $k\to\infty$.
% Generally, $k<1$ is considered as significant levels of overdispersion.
% Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocite{quiltyReconstructing,quiltyUnderstanding}.
% If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
% If it is Poisson distributed, then the standard deviation is on the order of the square root of the mean.
% A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
% The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.

@@

Both overdispersion and feedback loops violate the properties of a Poisson process.
Despite these issues, consistent inference on the incidence can still be produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.

@@

Now I introduce mechanistic models, in particular the class of mechanistic transmission models known as \emph{compartmental} models.
Compartmental models are the most widely-used class of mechanistic infectious disease model.
A compartmental model is named because they place every member of the population under study into a compartment; each compartment representing a disease state (\eg susceptible).
I will describe these models using the notation of the field, which differs in conventions from the rest of this thesis.
In particular, upper case or lower case will no longer be used to distinguish between random variables and their realisations, which instead is clear by context.

@@

Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters~\autocite{birrellEvidence,dukicTracking,corbellaThesis,keelingModeling}.
Stochastic models are most commonly specified in discrete time, using a Markov assumption.
A common simplification is to use discrete time and a Markov assumption; \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).

@@

The remainder of this section uses only deterministic models.
This choice is appropriate for the context of this thesis because there were many individuals infectious with SARS-CoV-2 in this time period.
Hence, further details on stochastic models are not outside the scope this thesis; a wealth of introductory texts with further details on stochastic models are available, I recommend \textcite[chapter 6]{keelingModeling}.

@@

Individuals in the susceptible compartment are those that could be infected (they have no immunity).
The proportion of the population that is susceptible at time $t$ is $s(t)$.
Individuals in the infectious compartment can infect others, and eventually recover.
The proportion of the population that is infectious at time $t$ is $i(t)$.
Individuals in the recovered compartment are immune to the disease, and cannot be infected again.
The proportion of the population that have recovered by time $t$ is $r(t)$.
The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.

@@

% The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
% Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infectious individual~\autocite[214]{kretzschmarMathematical}.
Assume this individual comes into contact with other individuals in the population at a mean rate of $\kappa$.
Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population; this is implied by assuming a homogeneous population.
For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Therefore the infectious individual makes contact with susceptible individuals at a mean rate of $\kappa s(t)$ per day.
Assume that there is a constant probability of transmission upon contact between an infectious and susceptible individual, $q$.
Hence, the mean rate of infections caused by this infectious individual is $\kappa s(t) q$ per day.
The two constant terms are absorbed into the transmission rate, $\beta = \kappa q$.
Therefore, the rate of new infections generated by a single infectious individual is $\beta s(t)$ per day.
Multiplying by the number of infectious individuals and dividing by the population size gives the rate per day at which susceptible individuals are infected as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.

@@

A few properties and assumptions of the SIR model are worth noting.
\begin{itemize}
    \item For any short period of time when the change in $s(t)$ is negligible, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
    A notable example is the early part of an epidemic when $s(t) \approx 1$.
    This is the phase of initial exponential growth of an epidemic, producing the characteristic early epidemic curves.
    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
    \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
    Intuitively, this can be thought of as the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
    \item The effective reproduction number for this model is $\Re = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
    $1-s(t)$ is the fraction of infections that are avoided because of the presence of immune individuals in the population.
    \item The population is assumed to be closed.
    This assumption implies several attributes.
    First, the total population size is constant.
    Second, only a negligible number of infections are imported from outside the population.
    Third, births and deaths of individuals are negligble over the length of time considered; model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
    \item There is no analytical solution to the system of ODEs: numerical methods are required to solve the system~\autocite[25]{keelingModeling}.
    \item As noted previously, the parameters $\vec{\theta}$ and the intial state of the system, $\vec{x}(0)$ are required to solve the system.
    \item Immunity is assumed to last until the end of the  period considered ~\autocite[61]{andersonInfectious}.
    In many contexts, including here, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
    For longer time periods, waning immunity can be incorporated by allowing individuals to move from the recovered compartment back to the susceptible compartment~\autocite[40]{keelingModeling}.
    \item The incidence proportion on day $t$ can be defined as $Z_t / \Npop = s(t) - s(t+1)$. However, this will not in general give an integer value to $Z_t$. In practice, this is not an issue since a continuous approximation for the incidence will suffice.
\end{itemize}

@@

The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
\todo{check this citation is formatted correctly}
This is done by adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
The ODEs are now:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
&\vdots \nonumber \\
\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
\frac{dr(t)}{dt} &= n\gamma i_n(t)
\end{align}
where $n$ is the number of infectious compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
This parameterisation is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.

@@

To relax the assumption of a homogeneous population, the population can be stratified.
If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
The different patterns could be due to behavioural factors (\eg high-numbers of sexual partners in a sexually-transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}.
I denote the set of strata, as $\set{A}$, with the set's size as $A$.
I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
Stratification requires two major  model changes: extension of the state vector and a description of the between-strata interactions.

The state vector is now a vector of vectors, one for each of the compartments
For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
The $a$th element of each vector ($a \in \set{A}$) is the proportion of strata $a$ in that compartment.
For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
These vectors replace the scalars in the previous equations.
Other, equivalent formulations of this model exist within the literature.
For example defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterised using numbers of individuals~\autocite[57]{keelingModeling}, rather than as the proportion of strata $a$ as I have done here.

@@

In practical situations, estimating all the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
It is infeasible because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
One common assumptions is symmetry, \ie $\beta_{aa'} = \beta_{a'a}$.
% Alternatively, as in the scalar case, it may be easier to iden $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
The choice of assumptions depends on the context, including the disease being studied and the data available.

>>>>

I start by defining these terms in \cref{inc-prev:sec:definitions}.
I then move on to explaining the three processes that are involved: the \emph{infection process}, the \emph{prevalence process}, and the \emph{observation process}.
The mathematical description of are an infection model, disease model, and observation model respectively.

@@

The prevalence process (\cref{inc-prev:sec:prevalence-process}) describes the relationship between incidence and prevalence.
Here, I adapt previous work to the SARS-CoV-2 context.
In particular, recovering from infection needs to be considered, which was not the case for HIV/AIDS.

@@

I use discrete time in this chapter, except in \cref{SEIR:sec:mechanistic-models}.
I use days; however, the results generalise in the obvious way to other time units.
I choose discrete time for pragmatic reasons.
Most epidemiological data is only available at daily granularity or, when more granular data exists, is often unreliable.
Furthermore, within-day patterns produce variations in transmission (\eg due to sleep) and data (\eg due to logistical considerations around collection); these variations are not of interest in most contexts.

\emph{Incidence} on day $t$, $Z_t$, is the number of people in a given population who become infected on day $t$; when expressed as a proportion of the population it is the \emph{incidence proportion}~\autocite[89]{lashModern}, $Z_t/\Npop$, where $\Npop$ is the number of people in the population.
The timing of infection events is very rarely directly observable and instead is inferred from other data.
Despite this issue, it remains the single most important quantity for informing the response to an epidemic or pandemic.

\emph{Prevalence} of a disease on day $t$, is the proportion of people with the disease on that day~\autocite[90]{lashModern}.
I refer to individuals with the disease as \emph{prevalent} individuals.
Whether an individual is prevalent is often expensive or error-prone to measure and hence is a latent quantity.
Furthermore, what is meant by ``having'' a disease can depend on context; for example, it could be the presence of symptoms, the ability to transmit, or detectable status (the difference between these was discussed in \cref{E-biology-data:sec:natural-history})
Within this thesis, I define an individual as prevalent on day $t$ if they are detectable (detectable is defined in \cref{E-biology-data:sec:detectable}).
I assume that an individual's prevalent status cannot change over the course of a day.
Therefore, prevalence is constant over any single day $t$.
I denote the number of prevalent individuals on day $t$ as $P_t$.
Therefore, the prevalence is $P_t/\Npop$.

\emph{Duration}, or \emph{time-to-event}, distributions are expressed in several equivalent ways~\autocite[17]{sunStatisticala}.
The duration distribution I will be concerned with is the distribution of the number of days for which an individual is prevalent.
Consider an individual $i$ infected on day $B_i$ that is prevalent during the interval $[B_i, E_i]$ (as in \cref{E-biology-data:sec:cis-episodes}).
I define this period as an \emph{infection episode}, beginning on day $B_i$ and ending on day $E_i$.
$D_i = E_i - B_i + 1$ is their \emph{duration of positivity}, the number of days for which they are prevalent.
I model the $D_i$s as a discrete independent and identically distributed (iid) random variable; extending the methodology to account for covariates is left to further work.
The duration is a positive integer, \ie $D_i \in \{1, 2, \dots, \dmax \}$, where $\dmax$ is the maximum duration.
The first way of expressing the distribution of $D_i$ is as the standard probability mass function (pmf), $f_{D}(d) = \prob(D = d)$; or cumulative density function (CDF), $F_{D}(d) = \prob(D \leq d) = \sum_{i=1}^d f_{D}(i)$.
The \emph{survival function} gives the probability that the duration is at least $d$ days: $S(d) = \prob(D \geq d) = 1 - F_{D}(D - 1)$ with $S(1) = 1$.
%(a common alternative convention for survival analysis with discrete variables is to define the survival function as $\prob(D >d)$).
The \emph{hazard} on day $d$ is the probability of recovering on day $d$ conditional on still being positive on day $d$: $\lambda(d) = \prob(D = d \mid D \geq d) = f_{D}(d) / S(d)$.
Throughout this thesis, the survival function will be the primary quantity of interest, and normally parameterized in terms of the hazard; these two quantities are related by $S(d) = \prod_{i=1}^{d-1} (1 - \lambda(i))$.

Since $D_i$ is a discrete random variable, $0 \leq f_D(d) \leq 1$ for all $d$ and $\sum_{k=1}^{\dmax} f_D(k) = 1$.
This implies that $F_D(d) \in [0, 1]$ and increases monotonically.
Therefore, $S(d) \in [0, 1]$ and decreases monotonically.
Similarly, $\lambda(d) \in [0,1]$, but can otherwise vary arbitrarily (\ie there is no monotonic or sum constraint).
The lack of monotonicity or sum constraint on the hazard makes it an attractive parameterization for inference.

@@

A point process is a random function of continuous time $K(t)$.
$K(t)$ must take non-negative integer values and be monotonically increasing in $t$.
The interpretation of $K(t)$ is the number of events, here infections, which have occurred in the interval $(-\infty, t]$~\autocite[244]{yanDistribution}.
In the context of a disease, the point process can be discretised into the incidence on day $t$ as $Z_t = K(t) - K(t-1)$, where $t$ is an integer.
I will explain two approaches to modelling the infection process.

First, in \cref{inc-prev:sec:poisson-process}, as a \emph{Poisson process}.
A Poisson process is a widely used type of counting process both within and outside of epidemiology.
It makes no direct assumptions on the underlying biological or epidemiological process and is therefore a flexible model.

@@

A commonly used point process for infectious disease epidemiology is the time inhomogeneous Poisson process~\autocite[e.g., in the context of HIV,][]{brookmeyerMethod,paganoHIV,rosenbergBackcalculation,brookmeyerBackcalculation}.
% A time-homogenous Poisson process is the special case where $\nu(r)$ is constant.
Extensions to this model, which I do not consider here, have incorporated covariates, such as geography or age~\autocite[e.g.][]{diggleModeling}.

@@

% A common cause of overdispersion is super-spreading events~\autocite{lloyd-smithSuperspreading}.
% A super-spreading event is an event where many more people than average are infected.
% That is, $Z_t >> \E(Z_t)$ for some $t$ during which a super-spreading event occurs.
% Coronaviruses, such as SARS-CoV-2, have particularly high overdispersion~\autocite{endoEstimating,adamClustering}{mccloskeySARS}.
% An infamous example of a SARS-CoV-2 super-spreading event was the Skagit County choir practice in March 2020~\autocite{hamnerHigh}.
% One infected individual is thought to have infected at least 32 further individuals during a single choir practice.
% For comparison, the average individual infected around three others in the early stages of the pandemic~\autocite{pellisChallenges}. 
% Overdispersion can be incorporated by replacing the assumption that $K(s) - K(t) \dist \Poi(\int_{t}^{s} \nu(r) dr)$ with $K(s) - K(t) \dist \NBc(\int_{t}^{s} \nu(r) dr, k)$ where $k$ is the overdispersion parameter, normally assumed to be constant absent interventions.
% Note that the expected number of events has not changed, but the variance has increased (see \cref{E-distributions} for details of parameterizations and moments).
% %The negative binomial tends to Poisson as $k\to\infty$.
% Generally, $k<1$ is considered as significant levels of overdispersion.
% Estimates for SARS-CoV-2 indicate $k \approx 0.1$~\autocite{endoEstimating}, and may be decreased by some interventions including lockdowns~\autocite{quiltyReconstructing,quiltyUnderstanding}.
% If $K(s) - K(t)$ has a negative binomial distribution, then its standard deviation scales on the order of the mean.
% If it is Poisson distributed, then the standard deviation is on the order of the square root of the mean.
% A particular implication of this is that the coefficient of variation (the standard deviation divided by the mean) tends to zero for a Poisson process, but tends to a constant for a negative binomial process.
% The disadvantage of the negative binomial distribution is that it loses much of the mathematical convenience of a Poisson distribution.

@@

Both overdispersion and feedback loops violate the properties of a Poisson process.
Despite these issues, consistent inference on the incidence is still produced using the Poisson process model, although the uncertainty in the estimates may be understated~\autocite{beckerDependent}.

@@

Now I will introduce mechanistic models, in particular the class of mechanistic transmission models known as \emph{compartmental} models.
Compartmental models are the most widely used class of mechanistic infectious disease models.
A compartmental model is named because they place every member of the population under study into a compartment; each compartment representing a disease state (\eg susceptible).
I will describe these models using the notation of the field, which differs in conventions from the rest of this thesis.
In particular, upper case or lower case will no longer be used to distinguish between random variables and their realisations, which instead is clear by context.

@@

Stochastic models specify a probability distribution for the current state $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta})$ where $\set{H}_t = \{ \vec{x}(t') \ssep t' < t \}$ is the epidemic's history and $\vec{\theta}$ are model parameters~\autocite{birrellEvidence,dukicTracking,corbellaThesis,keelingModeling}.
Stochastic models are most commonly specified in discrete time, using a Markov assumption.
A common simplification is to use discrete time and a Markov assumption, \ie $\vec{x}(t)$ is only defined for integer $t$ and $p(\vec{x}(t) \mid \set{H}_t, \vec{\theta}) = p(\vec{x}(t) \mid \vec{x}(t-1), \vec{\theta})$~\autocite{birrellEvidence}.
This relaxes the assumption of independent increments in the counting process formulation of the infection process (see \cref{E-inc-prev:sec:infection-process}).

@@

The remainder of this section uses only deterministic models.
This choice is appropriate for the context of this thesis because there were many individuals infectious with SARS-CoV-2 in this time period.
Hence, further details on stochastic models are outside the scope of this thesis; a wealth of introductory texts with further details on stochastic models are available, I recommend \textcite[chapter 6]{keelingModeling}.

@@

Individuals in the susceptible compartment are those that could be infected (they have no immunity).
The proportion of the population that is susceptible at time $t$ is $s(t)$.
Individuals in the infectious compartment can infect others, and eventually recover.
The proportion of the population that is infectious at time $t$ is $i(t)$.
Individuals in the recovered compartment are immune to the disease and cannot be infected again.
The proportion of the population that have recovered by time $t$ is $r(t)$.
The model's state is $\vec{x}(t) = (s(t), i(t), r(t))^T$.

@@

% The term $\beta s(t)i(t)$ is the rate at which susceptible individuals (as a proportion of the total population size) are infected~\autocite[18]{keelingModeling}.
% Susceptible individuals move from the $s(t)$ to $i(t)$ when they are infected.
The expression for the force of infection, $\lambda(t)$ can be arrived at by considering a single infectious individual~\autocite[214]{kretzschmarMathematical}.
Assume this individual comes into contact with other individuals in the population at a mean rate of $\kappa$.
Further, assume the population is \emph{well-mixed}, meaning that each contact is with an individual chosen uniformly at random from the population, implied by assuming a homogeneous population.
For such a randomly selected individual, the probability that they are susceptible is $s(t)$.
% Each of the infectious individual's contacts is therefore with a susceptible individual with probability $s(t)$, by the well-mixed assumption.
Therefore, the infectious individual makes contact with susceptible individuals at a mean rate of $\kappa s(t)$ per day.
Assume that there is a constant probability of transmission upon contact between an infectious and susceptible individual, $q$.
Hence, the mean rate of infections caused by this infectious individual is $\kappa s(t) q$ per day.
The transmission rate $\beta = \kappa q$ absorbs the two constant terms.
Therefore, $\beta s(t)$ is the rate of new infections generated by a single infectious individual per day.
Multiplying by the number of infectious individuals and dividing by the population size gives the daily rate of infection of susceptible individuals as a proportion of the population, $\beta s(t) i(t) = \lambda(t) s(t)$.

@@

A few properties and assumptions of the SIR model are worth noting.
\begin{itemize}
    \item For any short period of time when the change in $s(t)$ is negligible, the solution to the system of ODEs includes exponential growth in the number of infectious individuals: $i(t) = i_0 e^{\psi t}$ for some constants $i_0$ and $\psi$~\autocite[section 1.2]{diekmannMathematical}.
    A notable example is the early part of an epidemic when $s(t) \approx 1$.
    This is the phase of initial exponential growth of an epidemic, producing the characteristic early epidemic curves.
    It is in this phase that the basic reproduction number provides a good approximation of the epidemic dynamics.
    \item The basic reproduction number for the SIR model is $\R = \beta / \gamma$~\autocite[20]{keelingModeling}.
    This is the mean number of infections caused by an infectious individual per day, $\beta$, multiplied by the mean number of days for which they are infectious, $1/\gamma$.
    \item The effective reproduction number for this model is $\Re = \beta / \gamma s(t) = \R s(t)$~\autocite{pellisEstimation}.
    $1-s(t)$ is the fraction of infections that are avoided due to the presence of immune individuals in the population.
    \item The model assumes a closed population.
    This assumption implies several attributes.
    First, the total population size is constant.
    Second, imported infections are negligible. 
    Third, births and deaths of individuals are negligible over the length of time considered; model extensions allow these factors in contexts where they are relevant~\autocites[26]{keelingModeling}[214]{kretzschmarMathematical}.
    \item There is no analytical solution to the system of ODEs.
      Solving the system requires numerical methods~\autocite[25]{keelingModeling}.
    \item As noted previously, solving the system of ODEs requires the parameters $\vec{\theta}$ and the initial state of the system, $\vec{x}(0)$.
    \item The model assumes immunity lasts until the end of the period considered~\autocite[61]{andersonInfectious}.
    In many contexts, including here, this assumption is justified because the period is shorter than the length of immunity conferred~\autocite{milneImmunity}.
    For longer time periods, the model can be expanded to include waning immunity~\autocite[40]{keelingModeling}.
    \item The incidence proportion on day $t$ is defined as $Z_t / \Npop = s(t) - s(t+1)$. However, this will not in general give an integer value to $Z_t$. In practice, this is not an issue since a continuous approximation for the incidence will suffice.
\end{itemize}

@@

The basic SIR model is the deterministic approximation of a stochastic model where the duration of time in the infectious compartment is exponentially distributed~\autocite[96]{keelingModeling}.
This is an unrealistic assumption for most diseases and would lead to underestimating of the reproduction number~\autocite{lloydRealistic,wearingAppropriate}.
A standard extension to the SIR model is to instead assume a gamma distribution~\autocite[94]{keelingModeling,andersonSpread}.
\todo{check this citation is formatted correctly}
The extension adds adding multiple infectious compartments, shown in \cref{SEIR:fig:SIR-gamma}.
The ODEs are now:
\begin{align}
\frac{ds(t)}{dt} &= -\lambda(t) s(t)\\
\frac{di_1(t)}{dt} &= \lambda(t) s(t) - n\gamma i_1(t) \\
\frac{di_2(t)}{dt} &= n\gamma i_1(t) - n \gamma i_2(t) \\
&\vdots \nonumber \\
\frac{di_n(t)}{dt} &= n\gamma i_{n-1}(t) - n \gamma i_n(t) \\
\frac{dr(t)}{dt} &= n\gamma i_n(t)
\end{align}
where $n$ is the number of infectious compartments, and redefining $\lambda(t) = \beta \sum_{j=1}^n i_j(t)$.
The additional compartments added do not, in general, correspond to any biological feature or have a biological interpretation; they are purely a mathematical device to modify the distribution of time between becoming and ceasing to be infectious~\autocite{lloydRealistic}.
In this model the infectious period is distributed $\GamDist(n, n\gamma)$, a realistic model for a variety of diseases~\autocite{wearingAppropriate}.
This parameterization is chosen to give $1/\gamma$ as the mean infectious period (further details on the Gamma distribution are in \cref{E-distributions}).
A gamma distribution with the first parameter equal to 1 is an exponential distribution, in which case this model collapses to the basic SIR model in \cref{SEIR:sec:SIR}.
As $n$ increases, the between-individual variability in infectious period length reduces, tending to a constant as $n \to \infty$~\autocite{lloydRealistic}.

@@

To relax the assumption of a homogeneous population, the population can be stratified.
If the strata have differing transmission patterns, this can change the dynamics of the epidemic.
The different patterns could be due to behavioural factors (\eg high numbers of sexual partners in a sexually transmitted disease~\autocite[69]{keelingModeling} or younger age groups making many more contacts for respiratory diseases~\autocite[176]{andersonInfectious}), or biological factors such as genetics~\autocite[208]{andersonInfectious}.
I denote the set of strata, as $\set{A}$, with the set's size as $A$.
I will assume that the disease progression (\ie length of the exposed and infectious periods) is the same for all strata, although this assumption can be relaxed.
Stratification requires two major  model changes: extension of the state vector and a description of the between-strata interactions.

The state vector is now a vector of vectors, one for each of the compartments.
For example, in the basic SIR model, $\vec{x}(t) = (\vec{s}(t), \vec{i}(t), \vec{r}(t))^T$.
The $a$th element of each vector ($a \in \set{A}$) is the proportion of strata $a$ in that compartment.
For example, $\vec{s}(t) = (s_1(t), \dots, s_A(t))^T$ where $s_a(t)$ is the proportion of the $a$th strata that are susceptible at time $t$.
Hence, $N_a s_a(t)$ is the number of susceptible people in stratum $a$, where $N_a$ is the number of individuals in stratum $a$.
These vectors replace the scalars in the previous equations.
Other, equivalent formulations of this model exist within the literature.
For example, defining $s_a$ as the proportion of the total population that are both in strata $a$ and susceptible or parameterized using numbers of individuals~\autocite[57]{keelingModeling}, rather than as the proportion of strata $a$ as I have done here.

@@

In practical situations, estimating all the $\beta_{aa'}$ parameters is infeasible, so some assumptions must be made to reduce the number of parameters~\autocite[176]{andersonInfectious}.
It is infeasible because the progression of the epidemic depends on $\vec{\lambda}(t)$, which has dimension $A$, while $\matr{\beta}$ has dimension $A^2$.
A common assumption is symmetry, \ie $\beta_{aa'} = \beta_{a'a}$.
% Alternatively, as in the scalar case, it may be easier to iden $\beta_{aa'}$ can be decomposed into the contact rate between strata $a$ and $a'$, $\kappa_{aa'}$ and the probability of transmission upon contact, $q_{aa'}$, and then assumptions made about the components.
The choice of assumptions depends on the context, including the disease being studied and the data available.
